Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-07-15 supplement.tex
start adding supplementary results

@@

\subsection{PCA and \texttt{ADMIXTURE} analysis}

@@

For numerical convenience, observed genetic variants were binned by
recombination distances. We assessed the robustness of the statistics to errors
in the recombination maps by considering two different recombination maps, the
OMNI YRI and HapMapII
\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn}.
Statistics were largely unchanged by using a different map
(Figure~\ref{fig:supp-omni-hapmap-comparison}), although both maps were
inferred using array data, which is sparser than whole-genome sequencing data.
In subsequent analyses, we used data based on recombination rates from the OMNI YRI map.

@@

\subsection{Uncertainty and confidence intervals using Godambe methods}

@@

\section{Validations using simulations from inferred demographic models}

@@

\subsection{Simulation details}

@@

\subsection{cSFS prediction under inferred models}

@@

\section{Supplementary results}

>>>>

\aprcomment{TODO: expand this section to include further discussion about why
we chose each population, and why we have omitted some (such as those suggested
in the reviews).}

@@

\subsection{PCA and \texttt{ADMIXTURE} analysis}\label{sec:dimred}

@@

For numerical convenience, observed genetic variants were binned by
recombination distances. We assessed the robustness of the statistics to errors
in the recombination maps by considering three different recombination maps,
the OMNI YRI and HapMapII
\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn},
and a recombination map inferred from the Nama \citep{Van_Eeden2021-od}.
Statistics were largely unchanged by using a different map
(Figures~\ref{fig:supp-omni-hapmap-comparison},~\ref{fig:supp-omni-nama-comparison}).
The OMNI and HapMap recombination maps resulted in LD statistics that were
nearly identical, with only minor deviations between the OMNI/HapMap and the
Nama-inferred recombination maps. Both the OMNI and HapMap maps were inferred
using array data, which is sparser than whole-genome sequencing data as was
used in the Nama map inference. In subsequent analyses, we used data based on
recombination rates from the OMNI YRI map, and in the Supplemental Results we
show that the inferences are robust when using the Nama recombination map
(Section~\ref{sec:supp-alt-data}).

@@

\subsection{Uncertainty and confidence intervals}

@@

\section{Model validation}

We took two approaches to validate demographic model inferences. First, we
explored the effects of using alternative populations, sample sets, and
recombination maps, and the effect of relaxing fixed parameters when fitting
the demographic models to diversity and LD statistics. We then use simulations
to test our inferred model predictions against other informative summary
statistics. Results are shown in the Supplementary Results
(Section~\ref{sec:supp-results}).

\subsection{Validations using alternative data sets and model specification}

\subsubsection{Alternate Thousand Genomes population sets}

In our primary model fits, we used the Mende in Sierra Leone (MSL) and the
British from England and Scotland (GBR). We tested the same models against a
different West African population (the Yoruba from Ibidan, Nigeria, YRI), and
two different European populations (Western Europeans from Utah, CEU, and
Toscani in Italy, TSI). The CEU are closely related to GBR and are thus likely
to be an equivalently close proxy for European ancestry in African individuals
with admixed European ancestry. The TSI are likely more distantly related to
the source of this European ancestry. We refit each of our best-fit demographic
models using the following alternate population sets: (Nama, YRI, Gumuz,
Oromo/Amhara, and CEU), and (Nama, MSL, Gumuz, Oromo/Amhara, and TSI).

\subsubsection{Variable thresholds of Nama-associated ancestry}

The Nama have recent European ancestry, and Nama individuals in our dataset
show varying levels of inferred ancestry relating to European populations
(Figure~\ref{fig:diversity}D, in the main text). Each of our inferred models
allows for gene flow from Eurasia to Africa after the out-of-Africa migration
event, through early Holocene back-to-Africa migration in East Africa, recent
colonial admixture in South Africa, and ongoing continuous migration between
branches leading to East African and European populations. While our
demographic models directly account for European ancestry in Nama individuals,
inferred parameters relating to early divergences and population structure may
be impacted by recent admixture.

To assess the impact of recent European admixture in Nama individuals on
inferred early divergence times and population structure, we subset the set of
Nama individuals that we included based on inferred European-matching ancestry
proportions. We used inferred ancestry proportions from \texttt{ADMIXTURE} with
the number of clusters $K=4$ (Section~\ref{sec:dimred}) and kept individuals
with either at least a) 90\% or b) 99\% proportion of inferred ancestry that
clusters within the Nama group (colored green in Figure~\ref{fig:diversity}D).
This left a) 18 Nama individuals with at least 90\% inferred South African
ancestry, and b) 14 Nama individuals with at least 99\%. For each subset, we
recomputed diversity and heterozygosity statistics and fit each of our
demographic models to these statistics.

\subsubsection{Varying fixed divergence times}

In order to reduce parameter space to avoid overfitting, we fixed parameters
that have previously been inferred, including the timings of the early
back-to-Africa migration contributing to the ancestry of East African
agricultural populations (12ka), the divergence of West and East Africans
(60ka), and the out-of-Africa migration event (50ka). Due to variability in
inference and dating methods, there is some uncertainty surrounding these
dates. We explored how variable timings of these events effect our inferred
models by 1) increasing the fixed times of the E-W African split (to 100ka) and
the out-of-Africa event (to 60ka), and 2) allowing those two parameters to be
jointly fit in model optimization.

\subsubsection{Recombination map inferred from Nama}

Throughout, we used inferred recombination maps based on the YRI OMNI array
data \citep{1000_Genomes_Project_Consortium2015-zq}. Because recombination
rates can vary between populations, we tested the robustness of our demographic
inference to the recombination map used to compute LD. We recomputed diversity
and LD statistics using recombination rates from the Nama recombination map
\citep{Van_Eeden2021-od}, and refit our best-fit demographic models to these
statistics.

\subsubsection{\aprcomment{TODO:} High-coverage Nama samples}

\subsection{Validations using simulations from inferred demographic models}

@@

\subsubsection{Simulation details}

@@

\subsubsection{cSFS prediction under inferred models}

@@

\section{Supplementary results}\label{sec:supp-results}

\subsection{Model fit using alternative datasets}
\label{sec:supp-alt-data}

\begin{itemize}
    \item Using YRI and CEU, inferred parameters for each of our highlighted models
        did not change (all well within one SD of the inferred parameters of the
        original fit models)
    \item TODO: run fits with TSI
    \item Fits using the Nama recombination map, even though there were slight
        differences in some of the LD statistics (primarily a few noticeable
        differences in the $\pi_2$ statistics), also resulted in model fits that
        were very similar to the original fit models.
    \item Fits with alternate thresholds of Nama-inferred ancestry: TODO:
        fit continuous migration and the merger with stem migration model
\end{itemize}

\subsection{Relaxing fixed divergence times}

\begin{itemize}
    \item Fixed to older dates (100ka: E/W split, 60ka: OOA event). These fixed
        values stretched the other inferred parameters to older times, and resulted
        in a worse log-likelihood
    \item Allowed to vary, the best-fit inferred E/W split time and OOA event
        were more recent 35-50ka, instead of our fixed 50-60ka. \aprcomment{I
        still don't find this entirely unreasonable, as E Africans and Eurasian
        populations may be more closely related than the oldest evidence of H.
        sapiens in Europe (up to 50ka), or those earliest H. sapiens are likely
        not to be the direct ancestors of present-day Europeans. Still, it's
        probably too recent of a split, and fixing to 50ka is defensible.}
\end{itemize}
